---
title: "Cytometry analysis pipeline"
author: "Wlad Stroukov"
author: 
  - Wlad Stroukov,Immunoregulation laboratory, Peter Gorer Department of Immunology, School of Immunology and Microbial Science, King’s College London, wladislaw.stroukov@kcl.ac.uk
abstract: 
date: "09/10/2021"  
output:
  html_document: 
    code_folding: hide
    toc: true
    number_sections: true
    toc_float: true
---

```{r house_keeping, echo=TRUE}
## house keeping
rm(list = ls()) # clears the environment, in case something gets pre-loaded from a previous session
```

# References
Tutorial and Exercises by Christopher Hall. Link to [GitHub](https://github.com/hally166/Cytometry-R-scripts) and [Youtube](https://www.youtube.com/watch?v=2INqQNMNaV0)


# Data
Data can be accessed [here](https://flowrepository.org/id/FR-FCM-ZZZU)


# Installations
The following packages are required for the tutorial and subsequent data analysis.
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# 
# # Installing the "basic" flow cytometry packages.  They all live on a repository called Bioconductor.
# #install.packages("BiocManager")
# 
# #You can either do library(BiocManager) and install("flowCore) or what I have done below.
# BiocManager::install("flowCore") #interpret .fcs files
# BiocManager::install("flowViz") #basic visualisation
# BiocManager::install("ggcyto") #advanced visualisation using the ggPlot nomenclature
# BiocManager::install("openCyto") #Used to link various analysis methodologies
# BiocManager::install("flowWorkspace") #used to build anaylsis templates
# BiocManager::install("CytoML") #imports FlowJo and DiVA workspaces
# 
# #These packages largely require each other to work (except flowCore which is the "base package) 
# #so will often load each other without any help.
# 
# #Packages for cleaning data: flowClean, flowAI and flowCut (it's not necessary to have all three)
# #flowClean is the original, but is superseded by flowCut
# BiocManager::install("flowAI")
# BiocManager::install("flowClean")
# BiocManager::install("flowCut")
# #Some packages are not available on bioconductor and need to be loaded straight from GitHub.  TO do this you need the package devtools.
# install.packages("devtools")
# #An intresting project is CytoExploreR that trys to blend the power of R with the ease of use of a mouse.
# devtools::install_github("DillonHammill/CytoExploreRData")
# devtools::install_github("DillonHammill/CytoExploreR",build_vignettes = TRUE)
```


# Libraries
```{r}
library(flowCore) # Base package for Flow
library(flowAI)   # Data clean up and transformation
library(ggcyto)   # Visualisation
```


# Loading data
Loading a single file
```{r}
#getwd()
file <- "./Data/FlowRepository_FR-FCM-ZZZU_files/0001.FCS"
fcsfile <- read.FCS(file)
fcsfile
```
Loading multiple files
```{r}
files <- list.files(path="./Data/FlowRepository_FR-FCM-ZZZU_files", pattern=".FCS")
fs <- flowCore::read.flowSet(files[1:10], path="./Data/FlowRepository_FR-FCM-ZZZU_files/") # load first 10 files
fs[[1]]
```

# Compensation, Transformation, and Cleaning

Loading a file
```{r}
file <- "./Data/FlowRepository_FR-FCM-ZZZU_files/0004.FCS"
fcsfile <- read.FCS(file)
```


```{r}
fcsfile
#names(fcsfile) # list of parameters
#exprs(fcsfile) # list of expression data of all events/cells
#each_col(fcsfile,median) # applies function to each column
#keyword(fcsfile) # show meta data of file
keyword(fcsfile)$FILENAME
```
## Compensation
```{r}
spillover(fcsfile) # access the spill over matrix. In this file the matrix contains no data
fcsfile_comp <- compensate(fcsfile, spillover(fcsfile)$SPILL) # compensate the file with the spillovermatrix save in the file itself.
```

## Cleaning data
For a set of FCS files, flow_auto_qc performs a complete and automatic quality control. It consists in the detection and removal of anomalies by checking three properties of flow cytometry: 1) flow rate (instrument speed), 2) signal acquisition (signal intensity), 3) outside of dynamic range.
```{r, results="hide"}
fcsfile_comp_clean <- flow_auto_qc(fcsfile_comp) # from flowAI package

```


```{r}
fcsfile_comp_clean
```
#Transformation
Transforming data from a linear to log scale. Only fluorescence channels, not FSC and SSC
```{r}
transformed <- estimateLogicle(fcsfile_comp_clean, colnames(fcsfile_comp_clean[,3:10])) # generates calculations for transformation
fcsfile_comp_clean_trans <- transform(fcsfile_comp_clean, transformed)
```

# Visualisation
using ggcyto package  
plotting the transformed data
```{r message=TRUE, warning=TRUE}
#??ggcyto
#?autoplot
plot <- autoplot(fcsfile_comp_clean_trans) # automatic plotting
plot
```
Untransformed data (linear scale) as comparison
```{r}
autoplot(fcsfile_comp_clean)
```
```{r}
autoplot(fcsfile_comp_clean_trans, x="FITC-A", y="PE-A", bin=216) # bin argument changes the size of the hexagons
```
```{r}
autoplot(fcsfile_comp_clean_trans, x="time", y="PE-A", bin=200)
```
Visualisation of a flowset  
```{r}
files <- list.files(path="./Data/FlowRepository_FR-FCM-ZZZU_files", pattern=".FCS")
fs <- read.flowSet(files[1:10], path="./Data/FlowRepository_FR-FCM-ZZZU_files/")
#fs[[1]]
#spillover((fs[[4]]))
fs_comp <- compensate(fs, spillover(fs[[1]])$SPILL ) # this applies compensation of a single file to the whole set. 

fs_comp_clean <- flow_auto_qc(fs_comp)
trans <- estimateLogicle(fs_comp_clean[[1]], colnames(fs[[1]][,3:10]))
fs_comp_clean_trans <- transform(fs_comp_clean, trans)
autoplot(fs_comp_clean_trans, x="FITC-A")
```

using fsApply
```{r}
fsApply(fs, each_col, median) # alternative to each_col() funcion for single files
```


