---
title: "Cytometry analysis pipeline"
author: "Wlad Stroukov"
author: 
  - Wlad Stroukov,Immunoregulation laboratory, Peter Gorer Department of Immunology, School of Immunology and Microbial Science, King’s College London, wladislaw.stroukov@kcl.ac.uk
abstract: 
date: "09/10/2021"  
output:
  html_document: 
    code_folding: hide
    toc: true
    number_sections: true
    toc_float: true
---

```{r house_keeping, echo=TRUE}
## house keeping
rm(list = ls()) # clears the environment, in case something gets pre-loaded from a previous session
```

# References
Tutorial and Exercises by Christopher Hall. Link to [GitHub](https://github.com/hally166/Cytometry-R-scripts) and [Youtube](https://www.youtube.com/watch?v=2INqQNMNaV0)


# Data
Data can be accessed [here](https://flowrepository.org/id/FR-FCM-ZZZU)


# Installations
The following packages are required for the tutorial and subsequent data analysis.
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# 
# # Installing the "basic" flow cytometry packages.  They all live on a repository called Bioconductor.
# #install.packages("BiocManager")
# 
# #You can either do library(BiocManager) and install("flowCore) or what I have done below.
# BiocManager::install("flowCore") #interpret .fcs files
# BiocManager::install("flowViz") #basic visualisation
# BiocManager::install("ggcyto") #advanced visualisation using the ggPlot nomenclature
# BiocManager::install("openCyto") #Used to link various analysis methodologies
#BiocManager::install("flowWorkspace") #used to build anaylsis templates
# BiocManager::install("CytoML") #imports FlowJo and DiVA workspaces
# 
# #These packages largely require each other to work (except flowCore which is the "base package) 
# #so will often load each other without any help.
# 
# #Packages for cleaning data: flowClean, flowAI and flowCut (it's not necessary to have all three)
# #flowClean is the original, but is superseded by flowCut
# BiocManager::install("flowAI")
# BiocManager::install("flowClean")
# BiocManager::install("flowCut")
# #Some packages are not available on bioconductor and need to be loaded straight from GitHub.  TO do this you need the package devtools.
# install.packages("devtools")
# #An intresting project is CytoExploreR that trys to blend the power of R with the ease of use of a mouse.
# devtools::install_github("DillonHammill/CytoExploreRData")
# devtools::install_github("DillonHammill/CytoExploreR",build_vignettes = TRUE)
```


# Libraries
```{r}
library(flowCore) # Base package for Flow
library(flowAI)   # Data clean up and transformation
library(ggcyto)   # Visualisation
library(flowWorkspace) # gating, facilitate gating hiearchy similar to flowJo
library(openCyto)  # automated gating
library(gridExtra) # for vis of multiple plots
```


# Part I: Loading data
Loading a single file
```{r}
#getwd()
file <- "./Data/FlowRepository_FR-FCM-ZZZU_files/0001.FCS"
fcsfile <- read.FCS(file)
fcsfile
```
Loading multiple files
```{r}
files <- list.files(path="./Data/FlowRepository_FR-FCM-ZZZU_files", pattern=".FCS")
fs <- flowCore::read.flowSet(files[1:10], path="./Data/FlowRepository_FR-FCM-ZZZU_files/") # load first 10 files
fs
fs[[1]]
```

# Part II: Compensation, Transformation, and Cleaning

Loading a file
```{r}
file <- "./Data/FlowRepository_FR-FCM-ZZZU_files/0004.FCS"
fcsfile <- read.FCS(file)
```


```{r}
fcsfile
#names(fcsfile) # list of parameters
#exprs(fcsfile) # list of expression data of all events/cells
#each_col(fcsfile, median) # applies function to each column
#keyword(fcsfile) # show meta data of file
keyword(fcsfile)$FILENAME
```
## Compensation
```{r}
spillover(fcsfile) # access the spill over matrix. In this file the matrix contains no data. According to the file description the compensation is already applied.
fcsfile_comp <- compensate(fcsfile, spillover(fcsfile)$SPILL) # compensate the file with the spillovermatrix save in the file itself.
```

## Cleaning data
For a set of FCS files, flow_auto_qc performs a complete and automatic quality control. It consists in the detection and removal of anomalies by checking three properties of flow cytometry: 1) flow rate (instrument speed), 2) signal acquisition (signal intensity), 3) outside of dynamic range.
```{r, results="hide"}
fcsfile_comp_clean <- flow_auto_qc(fcsfile_comp, html_report = F) # from flowAI package, generate a html report. Use argument html_report = F to prevent that
#?flow_auto_qc
```

The cleaned file contains fewer cells compared to the raw file.
```{r}
fcsfile_comp_clean
```
#Transformation
Transforming data from a linear to log scale for better visualisation. Only fluorescence channels, not FSC and SSC
```{r}
transformed <- estimateLogicle(fcsfile_comp_clean, colnames(fcsfile_comp_clean[,3:10])) # generates calculations for transformation
fcsfile_comp_clean_trans <- transform(fcsfile_comp_clean, transformed)
```

# Visualisation
using ggcyto package  
plotting the transformed data  

Untransformed data (linear scale)
```{r message=TRUE, warning=TRUE}
#??ggcyto
#?autoplot
plot <- autoplot(fcsfile_comp_clean) # automatic plotting from ggcyto package
plot
```

transformed data (log scale)
```{r message=TRUE, warning=TRUE}
plot <- autoplot(fcsfile_comp_clean_trans) # automatic plotting
plot
```

```{r fig.height=4, fig.width=4}
autoplot(fcsfile_comp_clean_trans, x="FITC-A", y="PE-A", bin=256) # bin argument changes the size of the hexagons
```
```{r}
autoplot(fcsfile_comp_clean_trans, x="time", y="PE-A", bin=200)
```
Visualisation of a flowset  
```{r}
files <- list.files(path="./Data/FlowRepository_FR-FCM-ZZZU_files", pattern=".FCS")
fs <- read.flowSet(files[1:10], path="./Data/FlowRepository_FR-FCM-ZZZU_files/")
#fs[[1]]
#spillover((fs[[4]]))
fs_comp <- compensate(fs, spillover(fs[[1]])$SPILL ) # this applies compensation of a single file to the whole set. 

fs_comp_clean <- flow_auto_qc(fs_comp)
trans <- estimateLogicle(fs_comp_clean[[1]], colnames(fs[[1]][,3:10])) # perform the estimation using one file and select the columns using the same file
fs_comp_clean_trans <- transform(fs_comp_clean, trans) # apply the estimation to all files
autoplot(fs_comp_clean_trans, x="FITC-A")
```

using fsApply
```{r}
fsApply(fs, each_col, median) # equivalent to each_col() function for single files
```

# Part III: FlowWspace: Gating, Errors, Stats, Visualisations
Load necessary data and pre-process data
```{r}
files <- list.files(path="./Data/Treg/", pattern = ".fcs")
fs <- read.flowSet(files, path="./Data/Treg")
fs[[1]]
matrix <- spillover(fs[[1]])$SPILL
matrix

fs_comp <- compensate(fs, matrix)
fs_comp_clean <- flow_auto_qc(fs_comp)

names(fs_comp_clean[[1]])
trans <- estimateLogicle(fs_comp_clean[[1]], colnames(fs_comp_clean[,7:12])) # select the columns that should be transformed
fs_comp_clean_trans <- transform(fs_comp_clean, trans)

```

Visualisation
```{r}
fs_comp_clean_trans[[1]]
autoplot(fs_comp_clean_trans[[1]])
```
## Basic Gating
```{r fig.height=4, fig.width=4}
ggcyto(fs_comp_clean_trans[[1]], aes(x="FSC-A", y="SSC-A")) + geom_hex(bins=256)
```
Create empty gating set. The "first" gate is assigned `root` and acts as an "anchor". The initially defined gates have first to be applied to the root gate
```{r}
gs <- GatingSet(fs_comp_clean_trans)
```


Create gate for cells (flowWorkspace package)
```{r}
rg1 <- rectangleGate(filterId = "Lymphocytes", "FSC-A"=c(70000, 150000),"SSC-A"=c(0,70000) ) # define first gate and assign name
#?rectangleGate
gs_pop_add(gs, rg1, parent="root") # add gate to gatingset
recompute(gs) # recompute gatingset
```

Check gating set
```{r}
gs_get_pop_paths(gs)
```

```{r}
ggcyto(fs_comp_clean_trans[[1]], aes(x="FSC-A", y="SSC-A")) + 
  geom_hex(bins=256) + 
  geom_gate(gs_pop_get_gate(gs,"Lymphocytes"))
```

```{r}
gs_pop_get_stats(gs)
```
singlet gating  
```{r}
ggcyto(fs_comp_clean_trans[[1]], aes(x = "FSC-A", y = 'FSC-H'))+ geom_hex(bins = 256)

#?polygonGate
sqrcut <- matrix(c(50000,160000,160000,60000, 60000,160000,130000,50000),ncol=2,nrow=4)
colnames(sqrcut) <- c("FSC-A","FSC-H")
sqrcut
pg1 <- polygonGate(filterId="Singlets", .gate = sqrcut)


gs_pop_add(gs, pg1, parent = "Lymphocytes", name = "Singlets")
gs_get_pop_paths(gs)
recompute(gs)

ggcyto(fs_comp_clean_trans, aes(x = "FSC-A", y = 'FSC-H')) + 
  geom_hex(bins = 256)+ 
  geom_gate(gs_pop_get_gate(gs, "/Lymphocytes/Singlets"))
```

```{r}
plot(gs)
gs_pop_get_stats(gs)
gs_pop_get_stats(gs, "Singlets", "percent")
```


## Automatic gating
```{r}

```



